---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# osp.snapshots

<!-- badges: start -->

[![R-CMD-check](https://github.com/esqLABS/osp.snapshots/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/esqLABS/osp.snapshots/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/esqLABS/osp.snapshots/graph/badge.svg)](https://app.codecov.io/gh/esqLABS/osp.snapshots)
<!-- badges: end -->

The goal of osp.snapshots is to provide an R interface to work with PK-Sim project snapshots.

## Installation

You can install the development version of osp.snapshots from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("esqLABS/osp.snapshots")
```

## Basic Usage

### Importing a Snapshot

```{r, include=FALSE}
# Copy the test snapshot file to the package's extdata directory
file.copy(
  from = file.path("tests/testthat/data/test_snapshot.json"),
  to = file.path("inst", "extdata", "test_snapshot.json"),
  overwrite = TRUE
)
```

```{r, eval=FALSE}
library(osp.snapshots)

# Load a snapshot from a JSON file
snapshot <- load_snapshot("path/to/snapshot.json")
```

For this demo, we'll use a test snapshot included with the package:

```{r}
library(osp.snapshots)

# Load the test snapshot
snapshot_path <- system.file("extdata", "test_snapshot.json", package = "osp.snapshots")
snapshot <- load_snapshot(snapshot_path)

# View snapshot overview
snapshot
```

### Exploring Snapshot Contents

```{r}
# List all individuals in the snapshot
snapshot$individuals

# Get a detailed view of a building block
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`

# All fields are accessible as list elements
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$age
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$gender

# Parameters are also accessible as list element but they have their own nested structure and behavior
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$parameters

# They can be accessed by path
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$parameters$`Organism|Gallbladder|Gallbladder ejection fraction`
```


### Modifying Building Blocks
All these elements are mutable and can be modified in place.

```{r}
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$age <- 35

snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$parameters$`Organism|Gallbladder|Gallbladder ejection fraction`$value <- 0.6

snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`
```

## Creating new building blocks
### Creating a New Individual

```{r}
# Create a new individual
new_individual <- create_individual(
  name = "Patient_001",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "MALE",
  age = 45,
  weight = 85,
  height = 180
)

# Display the new individual
new_individual

# Add to snapshot
add_individual(snapshot, new_individual)

# Verify it was added
snapshot$individuals
```

### Creating a New Formulation

```{r}
# Create a simple dissolved formulation
dissolved_form <- create_formulation(
  name = "Oral Solution",
  type = "Dissolved"
)

# Create a tablet formulation with Weibull dissolution profile
tablet_form <- create_formulation(
  name = "Standard Tablet",
  type = "Weibull",
  parameters = list(
    dissolution_time = 60,
    dissolution_time_unit = "min",
    lag_time = 10,
    lag_time_unit = "min",
    dissolution_shape = 0.92,
    suspension = TRUE
  )
)

# Display the tablet formulation
tablet_form

# Add to snapshot
add_formulation(snapshot, tablet_form)

# Verify it was added
snapshot$formulations
```

## Converting to Data Frames

You can convert snapshot elements to data frames for easier analysis, visualization, and integration with other R tools.

### Converting Individuals

You can convert all individuals in a snapshot to combined data frames:

```{r,results='hide'}
# Get data frames for all individuals in the snapshot
dfs <- get_individuals_dfs(snapshot)
dfs
```

```{r, results='asis', echo=FALSE}
for (df in dfs) {
  print(knitr::kable(df))
}
```

It is also possible to get dataframe for a specific individual using `to_df()` method on an individual object:

```{r, results="hide"}
individual_df <- snapshot$individuals[[1]]$to_df()
individual_df
```
```{r, results='asis', echo=FALSE}
for (df in individual_df) {
  print(knitr::kable(df))
}
```

### Converting Formulations

Similar to individuals, you can convert formulations to data frames:

```{r, results="hide"}
# Get data frames for all formulations in the snapshot
formulation_dfs <- get_formulations_dfs(snapshot)
formulation_dfs
```

```{r, results='asis', echo=FALSE}
for (df in formulation_dfs) {
  print(knitr::kable(df))
}
```

You can also convert a specific formulation to a data frame:

```{r, results="hide"}
# Convert a specific formulation to data frame
form_df <- snapshot$formulations$`Standard Tablet`$to_df()
```

```{r, results='asis', echo=FALSE}
for (df in form_df) {
  print(knitr::kable(df))
}
```
