---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# osp.snapshots

<!-- badges: start -->

[![](https://img.shields.io/github/actions/workflow/status/esqlabs/osp.snapshots/merge-to-main.yaml?branch=main&label=Build)](https://github.com/esqlabs/osp.snapshots/actions/workflows/merge-to-main.yaml)
<!-- [![Codecov test coverage](https://codecov.io/gh/esqlabs/osp.snapshots/branch/main/graph/badge.svg)](https://app.codecov.io/gh/esqlabs/osp.snapshots?branch=main) -->

<!-- badges: end -->

osp.snapshots provides an R interface for working with PK-Sim project snapshots: import, navigate, modify, and export the nested JSON structure as R objects and data frames.

## Installation

You can install the development version of osp.snapshots from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("esqLABS/osp.snapshots")
```

## Usage

```{r example, message=FALSE}
library(osp.snapshots)

# Load a snapshot from a template name, a local .json path, or a URL
snapshot <- load_snapshot("Midazolam")
snapshot

# Navigate the building blocks
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`

# Convert a collection to tibbles for analysis
as_tibbles(snapshot, "individuals")
```

## Key features

* **Import and export**: Load snapshots from JSON files, URLs, or templates, and export the modified snapshot back to JSON.
* **Navigate**: Reach every building block (individuals, compounds, formulations, and the rest) with R syntax.
* **Modify**: Change parameters and add or remove building blocks in place.
* **Convert**: Turn any building block collection into structured tibbles with `as_tibbles()`.
* **Integrate**: Bridge PK-Sim observed data to `ospsuite::DataSet` objects.

## Learn more

* `vignette("osp-snapshots")` — getting started
* `vignette("creating-building-blocks")` — creating and managing building blocks
* `vignette("exporting-snapshots-to-dataframes")` — exporting snapshots to data frames
* `vignette("simulations")` — building simulations from scratch

## Getting help

* [Function reference](https://esqlabs.github.io/osp.snapshots/reference/)
* [Report bugs](https://github.com/esqLABS/osp.snapshots/issues)
