---
title: "Creating and managing building blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating and managing building blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
library(ospsuite)
```

## Overview

This vignette walks through the `create_*()` factory family, which lets you build PK-Sim *Building blocks* and their leaf sub-structures from named arguments instead of raw JSON. It also covers the mutators (`add_*()`, `remove_*()`) that attach these objects to a `Snapshot`.

We use the `Midazolam` template throughout:

```{r}
snapshot <- load_snapshot("Midazolam")
```

## Individuals

`create_individual()` builds an *Individual* from demographic arguments. PK-Sim seeds the underlying physiological tree at simulation time.

```{r}
patient <- create_individual(
  name = "Patient_001",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "FEMALE",
  age = 35,
  weight = 65,
  height = 165
)

patient
```

Extra demographics (e.g. `gestational_age`) are accepted as additional named arguments:

```{r}
pregnant_patient <- create_individual(
  name = "Pregnant_Patient",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "FEMALE",
  age = 28,
  weight = 70,
  height = 168,
  gestational_age = 20 # weeks
)

pregnant_patient
```

Attach the result to a snapshot with `add_individual()`:

```{r}
add_individual(snapshot, patient)
snapshot$individuals
```

## Formulations

`create_formulation()` covers every PK-Sim release profile (`"Dissolved"`, `"Weibull"`, `"First Order"`, `"Zero Order"`, `"Lint80"`, `"Particles"`, `"Table"`).

```{r}
oral_solution <- create_formulation(
  name = "Oral Solution 10mg/mL",
  type = "Dissolved"
)

oral_solution

tablet <- create_formulation(
  name = "Immediate Release Tablet",
  type = "Weibull",
  parameters = list(
    dissolution_time = 30,
    dissolution_time_unit = "min",
    lag_time = 5,
    lag_time_unit = "min",
    dissolution_shape = 1.2,
    suspension = TRUE
  )
)

tablet
```

```{r}
add_formulation(snapshot, oral_solution)
add_formulation(snapshot, tablet)
```

## Parameters

`create_parameter()` returns either a `Parameter` or a `LocalizedParameter` depending on the inputs. The two are kept separate because they identify themselves differently in the JSON: a plain `Parameter` lives inside a named container (e.g. a `Compound` or a `Formulation`) and carries a `Name`; a `LocalizedParameter` lives inside a parameter tree (typically an `Individual` or a `Simulation`) and carries a full pipe-separated `Path`.

`create_parameter()` routes on the `path` argument: passing `path = ...` produces a `LocalizedParameter`, passing only `name = ...` produces a `Parameter`. Both inherit from `Parameter`, so `inherits(x, "Parameter")` is `TRUE` for either.

```{r}
# Plain `Parameter` (name-based identity, suitable for a `Compound` or a
# `Formulation`)
clearance <- create_parameter(
  name = "Cl_spec",
  value = 5,
  unit = "ml/min/kg"
)
inherits(clearance, "Parameter")

# `LocalizedParameter` (path-based identity, suitable for an Individual's
# parameter tree)
ehc <- create_parameter(
  path = "Organism|Liver|EHC continuous fraction",
  value = 0.4
)
inherits(ehc, "LocalizedParameter")
```

Additional metadata for the *Value origin* (source, description, source identifier) is preserved through round-trip:

```{r}
gfr <- create_parameter(
  path = "Organism|Kidney|GFR",
  value = 120,
  unit = "ml/min",
  source = "Literature",
  description = "Normal adult GFR",
  source_id = 123
)

gfr
```

## Compounds

`create_compound()` wraps `Compound$new()` and validates common fields such as `molecular_weight_unit`:

```{r}
drug <- create_compound(name = "Drug X")

drug <- create_compound(
  name = "Drug X",
  is_small_molecule = TRUE,
  molecular_weight = 250.3,
  plasma_protein_binding_partner = "Albumin",
  parameters = list(
    create_parameter(name = "Cl_spec", value = 5, unit = "ml/min/kg")
  )
)

drug
```

### Compound processes

`create_process()` builds a *Process* (a PK-Sim `CompoundProcess`) from named arguments. The `internal_name` identifies the process template in PK-Sim's compound process repository (`"SpecificBinding"` for plasma protein binding, `"MetabolizationSpecific_MM"` for Michaelis-Menten metabolism, `"GlomerularFiltration"` for renal clearance, ...); `data_source` labels the data origin.

```{r}
binding <- create_process(
  internal_name = "SpecificBinding",
  data_source = "Buhr 1997",
  molecule = "GABRG2"
)

metabolism <- create_process(
  internal_name = "MetabolizationSpecific_MM",
  data_source = "Optimized",
  molecule = "CYP3A4",
  metabolite = "Hydroxy-Drug",
  parameters = list(
    create_parameter(name = "Km", value = 1.2, unit = "µmol/l"),
    create_parameter(name = "kcat", value = 9.865, unit = "1/min")
  )
)

metabolism$category
```

The `$category` active binding derives a domain label (`"protein_binding_partners"`, `"metabolizing_enzymes"`, `"hepatic_clearance"`, `"transporter_proteins"`, `"renal_clearance"`, `"biliary_clearance"`, `"inhibition"`, or `"induction"`) from the `internal_name`. Compound processes loaded from a snapshot are exposed as a flat named list of *Process* objects via `compound$processes`, and the long-form `processes` tibble returned by `get_compounds_dfs()` surfaces every (process, parameter) pair as a row.

## Populations

`create_population()` builds a *Population* recipe: settings used by PK-Sim to sample a cohort at simulation time. Use `range()` for age, weight, height, and BMI bounds.

```{r}
adults <- create_population(
  name = "Healthy Adults",
  number_of_individuals = 50,
  proportion_of_females = 50,
  species = "Human",
  source_population = "European_ICRP_2002",
  age_range = range(20, 60, "year(s)"),
  weight_range = range(50, 90, "kg")
)

adults
```

## Expression profiles

`create_expression_profile()` builds an *ExpressionProfile*. The identity of a profile is the composite `Molecule|Species|Category`, so all three are required, along with the molecule `type`.

```{r}
cyp3a4 <- create_expression_profile(
  molecule = "CYP3A4",
  species = "Human",
  category = "Healthy",
  type = "Enzyme",
  ontogeny = "CYP3A4"
)

cyp3a4
```

## Protocols

`create_protocol()` builds a *Protocol*. By default it creates a *Simple Protocol* (one application type, one dose, optional dosing interval); pass `schemas` to create an *Advanced Protocol*.

```{r}
single_dose <- create_protocol(
  name = "Single dose 10mg",
  application_type = "Oral",
  dosing_interval = "Single",
  parameters = list(
    create_parameter(name = "Start time", value = 0, unit = "h"),
    create_parameter(name = "InputDose", value = 10, unit = "mg")
  )
)

single_dose
```

### Advanced protocols

For dosing schedules that do not fit the *Simple Protocol* pattern, build an *Advanced Protocol* from one or more *Schema* blocks. Each *Schema* owns schema-level parameters (such as `NumberOfRepetitions` and `TimeBetweenRepetitions`) and an ordered list of *SchemaItem* applications.

`create_schema()` and `create_schema_item()` wrap the R6 constructors so you can pass *Parameter* objects directly rather than hand-rolling the raw JSON shape. We recommend the factory form for new code; the raw-list form remains accepted for forward compatibility with hand-edited JSON.

```{r}
qd_oral <- create_protocol(
  name = "Once daily oral, 5 days",
  schemas = list(
    create_schema(
      name = "Schema 1",
      parameters = list(
        create_parameter(name = "NumberOfRepetitions", value = 5),
        create_parameter(name = "TimeBetweenRepetitions", value = 24, unit = "h")
      ),
      items = list(
        create_schema_item(
          name = "Item 1",
          application_type = "Oral",
          parameters = list(
            create_parameter(name = "Start time", value = 0, unit = "h"),
            create_parameter(name = "InputDose", value = 10, unit = "mg")
          )
        )
      )
    )
  ),
  time_unit = "h"
)

qd_oral
```

`application_type` on a schema item is validated against the canonical PK-Sim application types (`"Oral"`, `"IntravenousBolus"`, `"IntravenousInfusion"`, `"Intramuscular"`, `"Subcutaneous"`, `"Dermal"`, `"Rectal"`, `"Inhalation"`, `"Intraperitoneal"`).

## Events

`create_event()` builds an *Event* from a named template (for example a meal). PK-Sim clones the template at simulation time and applies the parameter overrides supplied here.

```{r}
breakfast <- create_event(
  name = "Breakfast",
  template = "Meal: Standard (Human)",
  parameters = list(
    create_parameter(name = "Meal energy content", value = 500, unit = "kcal"),
    create_parameter(name = "Meal volume", value = 0.3, unit = "l")
  )
)

breakfast
```

## Observer sets

`create_observer_set()` builds an *ObserverSet* from a name and an optional list of observers. Each *Observer* is a simulation-time formula that exposes a derived quantity (for example a tissue concentration) that is not a natural model output.

The factory accepts either *Observer* R6 objects (recommended) or raw observer lists. Prefer the R6 form: `Observer$new()` validates the input and gives you the typed accessors (`$name`, `$type`, `$dimension`, `$formula`, `$container_tags`).

```{r}
# Recommended: build observers with `Observer$new()` and pass them in
observer <- Observer$new(list(
  Name = "brain_plasma_conc",
  Type = "Container",
  Dimension = "Concentration (molar)",
  Formula = list(Formula = "Conc_Br")
))

brain_set <- create_observer_set(
  name = "BrainPlasmaConcentration",
  observers = list(observer)
)

brain_set

# Still accepted for round-tripping hand-edited JSON: raw observer lists
brain_set_raw <- create_observer_set(
  name = "BrainPlasmaConcentration",
  observers = list(
    list(
      Name = "brain_plasma_conc",
      Type = "Container",
      Dimension = "Concentration (molar)",
      Formula = list(Formula = "Conc_Br")
    )
  )
)

brain_set_raw
```

## Observed data

`create_observed_data()` builds an `ospsuite::DataSet` from named arguments for the time grid, measurement values, units, and optional error series. `value_dimension` is required and gates the unit validation.

```{r}
obs <- create_observed_data(
  name = "Phase I - Subject 001",
  time = c(0, 0.5, 1, 2, 4, 8, 12, 24),
  values = c(0, 12.5, 18.2, 15.8, 11.2, 6.8, 3.4, 1.1),
  time_unit = "h",
  value_unit = "mg/l",
  value_dimension = "Concentration (mass)"
)

obs
```

## Managing building blocks

Each building-block kind has paired `add_*()` and `remove_*()` mutators that take the snapshot first and return it invisibly, so they chain with the base pipe.

```{r}
remove_individual(snapshot, "Patient_001")
remove_formulation(snapshot, "Oral Solution 10mg/mL")
```

You can pass several names to remove at once:

```{r, eval=FALSE}
remove_individual(snapshot, c("Patient_001", "Patient_002"))
remove_expression_profile(snapshot, "CYP3A4|Human|Healthy")
```

### Batch creation

```{r}
patients <- list(
  create_individual("Patient_A", age = 25, gender = "MALE"),
  create_individual("Patient_B", age = 45, gender = "FEMALE"),
  create_individual("Patient_C", age = 65, gender = "MALE")
)

for (patient in patients) {
  add_individual(snapshot, patient)
}

snapshot$individuals
```

### Modifying existing building blocks

Fields on existing building blocks are mutable through active bindings:

```{r}
patient <- snapshot$individuals$Patient_A
patient$age <- 30
patient$weight <- 75
```

### Validation

The factories validate common inputs and abort with informative messages on bad input:

```{r, error=TRUE}
tryCatch(
  {
    create_individual(
      name = "Invalid Patient",
      species = "InvalidSpecies",
      gender = "OTHER"
    )
  },
  error = function(e) cat("Error:", e$message, "\n")
)

tryCatch(
  {
    create_formulation(
      name = "Invalid Form",
      type = "NonexistentType"
    )
  },
  error = function(e) cat("Error:", e$message, "\n")
)
```

The validation helpers (`validate_species()`, `validate_gender()`, `validate_unit()`, `validate_population()`) check inputs against the canonical ospsuite enums and are exported for direct use.

## Exporting

After your modifications, write the snapshot back to JSON:

```{r, eval=FALSE}
export_snapshot(snapshot, "modified_snapshot.json")
```

## Next steps

* `vignette("working-with-dataframes")` covers the tibble layer for analysis.
* The full API reference is at <https://esqlabs.github.io/osp.snapshots/reference/>.
