---
title: "Creating and managing building blocks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating and managing building blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
library(ospsuite)
```

## Overview

This vignette walks through the `create_*()` functions that build PK-Sim *Building blocks* from named arguments, and the `add_*()` / `remove_*()` mutators that attach them to a snapshot.

We use the `Midazolam` template throughout:

```{r}
snapshot <- load_snapshot("Midazolam")
```

## Starting from an empty snapshot

`create_snapshot()` is the from-scratch counterpart to `load_snapshot()`. It returns an in-memory *Snapshot* that carries the current supported PK-Sim version and no *Building blocks*, touches no files, and has no path. You then populate it with the `add_*()` verbs and serialize it with `export_snapshot()`.

```{r}
empty <- create_snapshot(name = "My Project", description = "Notes")
empty <- add_compound(empty, create_compound(name = "Drug X"))
empty
```

The rest of this vignette mutates the `Midazolam`-based `snapshot` object, so the empty snapshot uses its own `empty` object and does not disturb that narrative.

## Individuals

`create_individual()` builds an *Individual* from demographic arguments. PK-Sim seeds the underlying physiological tree at simulation time.

```{r}
patient <- create_individual(
  name = "Patient_001",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "FEMALE",
  age = age(35),
  weight = weight(65),
  height = height(165)
)

patient
```

Each demographic field takes a value-object helper (`age()`, `weight()`, `height()`, `gestational_age()`) that carries the value and its unit; the helper owns the default unit and validates a supplied unit. Extra demographics (e.g. `gestational_age`) are accepted the same way:

```{r}
pregnant_patient <- create_individual(
  name = "Pregnant_Patient",
  species = "Human",
  population = "European_ICRP_2002",
  gender = "FEMALE",
  age = age(28),
  weight = weight(70),
  height = height(168),
  gestational_age = gestational_age(20) # weeks
)

pregnant_patient
```

Attach the result to a snapshot with `add_individual()`:

```{r}
snapshot <- add_individual(snapshot, patient)
snapshot$individuals
```

Parameters that sit on a specific path inside the individual's physiological tree are built with `create_parameter()` using a `path` argument. PK-Sim writes the path as pipe-separated segments (`Organism|Liver|...`):

```{r}
ehc <- create_parameter(
  path = "Organism|Liver|EHC continuous fraction",
  value = 0.4
)

gfr <- create_parameter(
  path = "Organism|Kidney|GFR",
  value = 120,
  unit = "ml/min",
  source = "Literature",
  description = "Normal adult GFR",
  source_id = 123
)

gfr
```

You can pass a list of such parameters straight to `create_individual(parameters = ...)`, or assign them through `individual$parameters` after the fact (see "Modifying existing building blocks" below).

## Formulations

`create_formulation()` builds a *Formulation* from a `type` and its `parameters`. The `type` argument is open rather than a closed set. Seven curated human aliases resolve to a known PK-Sim release profile and unlock a per-type curated parameter vocabulary: `"Dissolved"`, `"Weibull"`, `"Lint80"`, `"Particle"`, `"Table"`, `"Zero Order"`, and `"First Order"`. Any other non-empty string is accepted verbatim and written to `FormulationType`, so you can author a new or unknown PK-Sim template type the same way `create_event()` accepts an arbitrary `template`.

The `parameters` argument accepts two mutually exclusive forms. The curated alias form takes bare scalar or vector values keyed by the per-type alias vocabulary, and it exists only for the seven known types. For example, `"Dissolved"` takes none, and `"Weibull"` accepts `dissolution_time`, `dissolution_shape`, `lag_time`, and related keys. See `?create_formulation` for the full parameter list per type. The raw form takes a list of `create_parameter()` objects (or raw `list(Name =, Value =, ...)` dicts) written straight to the formulation's `Parameters`, preserving `Unit`, `ValueOrigin`, and `TableFormula` verbatim, for any type. The raw form is the only valid form for an unknown type.

```{r}
oral_solution <- create_formulation(
  name = "Oral Solution 10mg/mL",
  type = "Dissolved"
)

oral_solution

tablet <- create_formulation(
  name = "Immediate Release Tablet",
  type = "Weibull",
  parameters = list(
    dissolution_time = 30,
    dissolution_time_unit = "min",
    lag_time = 5,
    lag_time_unit = "min",
    dissolution_shape = 1.2,
    suspension = TRUE
  )
)

tablet
```

The raw form lets you set an arbitrary parameter `Name`, a per-parameter `ValueOrigin` (`source` and `description`), or a custom `TableFormula` on any type, including `"Dissolved"` and unknown types:

```{r}
raw_dissolved <- create_formulation(
  name = "Suspension",
  type = "Dissolved",
  parameters = list(
    create_parameter(
      name = "Use as suspension",
      value = 1,
      source = "Lit",
      description = "Reference XYZ"
    )
  )
)

raw_dissolved
```

```{r}
snapshot <- add_formulation(snapshot, oral_solution)
snapshot <- add_formulation(snapshot, tablet)
```

## Compounds

`create_compound()` builds a compound with validation on common fields such as `molecular_weight_unit`. Compound-level parameters (clearances, halogens, ...) are passed as a list of `create_parameter()` results: parameters named on a compound are identified by `name`, so omit the `path` argument.

```{r}
drug <- create_compound(name = "Drug X")

drug <- create_compound(
  name = "Drug X",
  is_small_molecule = TRUE,
  molecular_weight = 250.3,
  plasma_protein_binding_partner = "Albumin",
  parameters = list(
    create_parameter(name = "Cl_spec", value = 5, unit = "ml/min/kg")
  )
)

drug
```

### Compound processes

`create_process()` builds a compound process (plasma protein binding, metabolism, clearance, ...) from named arguments. The `internal_name` identifies the process template in PK-Sim's compound process repository (for example `"SpecificBinding"`, `"MetabolizationSpecific_MM"`, `"GlomerularFiltration"`); `data_source` labels the data origin. See `?create_process` for usage and current examples; the canonical list of `internal_name` values is maintained in PK-Sim and groups into the categories surfaced by `process$category` (`protein_binding_partners`, `metabolizing_enzymes`, `hepatic_clearance`, `transporter_proteins`, `renal_clearance`, `biliary_clearance`, `inhibition`, `induction`).

```{r}
binding <- create_process(
  internal_name = "SpecificBinding",
  data_source = "Buhr 1997",
  molecule = "GABRG2"
)

metabolism <- create_process(
  internal_name = "MetabolizationSpecific_MM",
  data_source = "Optimized",
  molecule = "CYP3A4",
  metabolite = "Hydroxy-Drug",
  parameters = list(
    create_parameter(name = "Km", value = 1.2, unit = "µmol/l"),
    create_parameter(name = "kcat", value = 9.865, unit = "1/min")
  )
)

metabolism$category
```

The `$category` active binding derives a domain label (`"protein_binding_partners"`, `"metabolizing_enzymes"`, `"hepatic_clearance"`, `"transporter_proteins"`, `"renal_clearance"`, `"biliary_clearance"`, `"inhibition"`, or `"induction"`) from the `internal_name`. Compound processes loaded from a snapshot are exposed as a flat named list of *Process* objects via `compound$processes`, and the long-form `processes` tibble returned by `get_compounds_dfs()` surfaces every (process, parameter) pair as a row.

## Populations

`create_population()` builds a *Population* recipe: settings used by PK-Sim to sample a cohort at simulation time. Use `range()` for age, weight, height, and BMI bounds.

```{r}
adults <- create_population(
  name = "Healthy Adults",
  number_of_individuals = 50,
  proportion_of_females = 50,
  species = "Human",
  source_population = "European_ICRP_2002",
  age_range = range(20, 60, "year(s)"),
  weight_range = range(50, 90, "kg")
)

adults
```

## Expression profiles

`create_expression_profile()` builds an *ExpressionProfile*. The identity of a profile is the composite `Molecule|Species|Category`, so all three are required, along with the molecule `type`.

```{r}
cyp3a4 <- create_expression_profile(
  molecule = "CYP3A4",
  species = "Human",
  category = "Healthy",
  type = "Enzyme",
  ontogeny = "CYP3A4"
)

cyp3a4
```

## Protocols

`create_protocol()` builds a *Protocol*. By default it creates a *Simple Protocol* (one application type, one dose, optional dosing interval); pass `schemas` to create an *Advanced Protocol*.

```{r}
single_dose <- create_protocol(
  name = "Single dose 10mg",
  application_type = "Oral",
  dosing_interval = "Single",
  parameters = list(
    create_parameter(name = "Start time", value = 0, unit = "h"),
    create_parameter(name = "InputDose", value = 10, unit = "mg")
  )
)

single_dose
```

### Advanced protocols

For dosing schedules that do not fit a single application (a multi-day course, or two compounds dosed on staggered schedules), use an *Advanced Protocol*. An Advanced Protocol is built from one or more *Schemas*; each schema describes a repeating block of applications (for example "10 mg orally every 24 hours for 5 days"). Inside a schema, each individual application is a *SchemaItem*.

You build them with `create_schema()` (the repeating block) and `create_schema_item()` (one application), then pass a list of schemas to `create_protocol()`:

```{r}
qd_oral <- create_protocol(
  name = "Once daily oral, 5 days",
  schemas = list(
    create_schema(
      name = "Schema 1",
      parameters = list(
        create_parameter(name = "NumberOfRepetitions", value = 5),
        create_parameter(name = "TimeBetweenRepetitions", value = 24, unit = "h")
      ),
      items = list(
        create_schema_item(
          name = "Item 1",
          application_type = "Oral",
          parameters = list(
            create_parameter(name = "Start time", value = 0, unit = "h"),
            create_parameter(name = "InputDose", value = 10, unit = "mg")
          )
        )
      )
    )
  ),
  time_unit = "h"
)

qd_oral
```

`application_type` on a schema item is validated against the canonical PK-Sim application types (`"Oral"`, `"IntravenousBolus"`, `"IntravenousInfusion"`, `"Intramuscular"`, `"Subcutaneous"`, `"Dermal"`, `"Rectal"`, `"Inhalation"`, `"Intraperitoneal"`).

## Events

An *Event* is a discrete perturbation that is not a drug administration: a meal, a change of posture, an enzyme inducer dose, an exercise bout. `create_event()` builds one from a `template` (the name of an event recipe in PK-Sim's event database) plus optional parameter overrides. PK-Sim clones the named template at simulation time and applies your overrides.

The set of available templates is defined inside PK-Sim (in particular by the PK-Sim Events database that ships with the installation), so the exact list of valid `template` names depends on your PK-Sim setup. Templates that are commonly seen in published snapshots include the "Meal: Standard (Human)" family for food effects; if you need a template that is not on your installation, define it in PK-Sim first and re-export.

```{r}
breakfast <- create_event(
  name = "Breakfast",
  template = "Meal: Standard (Human)",
  parameters = list(
    create_parameter(name = "Meal energy content", value = 500, unit = "kcal"),
    create_parameter(name = "Meal volume", value = 0.3, unit = "l")
  )
)

breakfast
```

## Observer sets

An *Observer* is a simulation-time formula that exposes a derived quantity (for example a tissue concentration) that is not a natural model output. Observers travel in groups called *Observer sets*. `create_observer_set()` builds a set from a name and a list of observers.

Build each observer with `create_observer()`. It validates the input and returns an *Observer* with typed accessors `$name`, `$type`, `$dimension`, `$formula` (the full `ExplicitFormula` list), `$formula_expression`, `$formula_dimension`, `$formula_references`, and `$container_tags`. The companions `create_formula_reference()`, `create_descriptor_condition()`, and `create_molecule_list()` build the observer's sub-properties. Bundle the observers into a set and attach the set to a snapshot:

```{r}
brain_obs <- create_observer(
  name = "brain_plasma_conc",
  type = "Container",
  dimension = "Concentration (molar)",
  formula = "Conc_Br",
  formula_references = list(
    create_formula_reference(
      "Conc_Br",
      "Organism|Brain|Plasma|Midazolam|Concentration"
    )
  ),
  container_criteria = list(
    create_descriptor_condition("Brain", "InContainer")
  ),
  molecule_list = create_molecule_list(for_all = FALSE, include = "Midazolam")
)

brain_set <- create_observer_set(
  name = "BrainPlasmaConcentration",
  observers = list(brain_obs)
)

brain_set

snapshot <- add_observer_set(snapshot, brain_set)
```

## Observed data

`create_observed_data()` builds an `ospsuite::DataSet` from value-object helpers for the time grid (`time()`), the measurement values (`values()`), and an optional error series (`error()`). The values dimension is required and is supplied to `values()`; it gates the unit validation. `time()` defaults its unit to `"h"`, and `error`, `molecular_weight`, `lloq`, and `metadata` are optional.

```{r}
obs <- create_observed_data(
  name = "Phase I - Subject 001",
  time = time(c(0, 0.5, 1, 2, 4, 8, 12, 24), unit = "h"),
  values = values(
    c(0, 12.5, 18.2, 15.8, 11.2, 6.8, 3.4, 1.1),
    unit = "mg/l",
    dimension = "Concentration (mass)"
  )
)

obs
```

`create_observed_data()` is a thin factory around `loadDataSetFromSnapshot()`, the lower-level function that converts a raw snapshot observed-data structure into an `ospsuite::DataSet`. Call `loadDataSetFromSnapshot()` directly when you already hold the raw structure (for example a slice read from a snapshot JSON) rather than the named-argument inputs.

The bridge also exposes two time helpers. `convert_ospsuite_time_unit_to_lubridate()` maps an ospsuite time unit to its lubridate unit, and `convert_ospsuite_time_to_duration()` maps a value and unit to a lubridate duration. The `"ks"` (kilosecond) unit is a special case: the duration helper converts it to seconds by multiplying the value by 1000.

```{r}
convert_ospsuite_time_unit_to_lubridate("day(s)")
convert_ospsuite_time_to_duration(2, "ks")
```

<div class="alert alert-warning">
The observed-data export path is asymmetric. A loaded observed-data entry replays byte-for-byte from the original JSON on export, so post-load mutations of a loaded `DataSet` are not reflected on export. A `DataSet` added at runtime with `add_observed_data()` is serialized through a lossy adapter that round-trips through `osp.snapshots` itself, but its PK-Sim round-trip is unvalidated. Treat runtime-added observed data as a convenience for the R side, not as a guaranteed PK-Sim round-trip.
</div>

## Managing building blocks

Each building-block kind has paired `add_*()` and `remove_*()` mutators that take the snapshot first and return it invisibly, so they chain with the base pipe.

```{r}
snapshot <- remove_individual(snapshot, "Patient_001")
snapshot <- remove_formulation(snapshot, "Oral Solution 10mg/mL")
```

You can pass several names to remove at once:

```{r, eval=FALSE}
snapshot <- remove_individual(snapshot, c("Patient_001", "Patient_002"))
snapshot <- remove_expression_profile(snapshot, "CYP3A4|Human|Healthy")
```

### Adding several building blocks at once

`add_*()` accepts either a single building block or a list of them, so you can attach many in one call. Each `add_*()` reports how many entries were added.

```{r}
patients <- list(
  create_individual("Patient_A", age = age(25), gender = "MALE"),
  create_individual("Patient_B", age = age(45), gender = "FEMALE"),
  create_individual("Patient_C", age = age(65), gender = "MALE")
)

snapshot <- add_individual(snapshot, patients)

snapshot$individuals
```

### Modifying existing building blocks

Most fields on an existing building block are mutable; assign to them like any other R6 active binding. Reading the object before and after shows the effect:

```{r}
patient <- snapshot$individuals$Patient_A
patient
```

```{r}
patient$age <- age(30)
patient$weight <- weight(75)
patient$gender <- "FEMALE"
patient
```

The same pattern works for nested objects. To bump a parameter on an individual, look it up by path under `individual$parameters` and write to its `value`:

```{r}
korean <- snapshot$individuals$`Korean (Yu 2004 study)`
ehc <- korean$parameters$`Organism|Liver|EHC continuous fraction`
ehc
ehc$value <- 0.4
ehc
```

After every change, the parent snapshot exports the updated values: `export_snapshot(snapshot, ...)` writes them straight back to JSON, no extra step required.

### Validation

`create_*()` functions validate common inputs against the canonical ospsuite enums and abort with informative messages on bad input:

```{r, error=TRUE}
tryCatch(
  {
    create_individual(
      name = "Invalid Patient",
      species = "InvalidSpecies",
      gender = "OTHER"
    )
  },
  error = function(e) cat("Error:", e$message, "\n")
)

tryCatch(
  {
    create_formulation(
      name = "Invalid Form",
      type = "NonexistentType"
    )
  },
  error = function(e) cat("Error:", e$message, "\n")
)
```

The same checks are exported as standalone helpers when you need them outside a `create_*()` call. Each returns `TRUE` on success and aborts with an informative message otherwise:

* `validate_species()` checks a species against `ospsuite::Species`.
* `validate_gender()` checks a gender against `ospsuite::Gender`.
* `validate_population()` checks a population against `ospsuite::HumanPopulation`, the human population enum, not a general population list.
* `validate_unit()` checks a unit against a dimension via ospsuite.
* `validate_snapshot()` checks that an object is a *Snapshot*.

```{r}
validate_species("Human")
validate_unit("mg/l", "Concentration (mass)")
```

## Exporting

After your modifications, write the snapshot back to JSON:

```{r, eval=FALSE}
export_snapshot(snapshot, "modified_snapshot.json")
```

## Next steps

* `vignette("exporting-snapshots-to-dataframes")` covers the tibble layer for analysis.
* The full API reference is at <https://esqlabs.github.io/osp.snapshots/reference/>.
