---
title: "Exporting snapshots to data frames"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting snapshots to data frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
library(dplyr)
```

## Overview

osp.snapshots can flatten the nested PK-Sim snapshot tree into tibbles ready for downstream analysis with `dplyr`. This vignette covers the entry point, the per-kind shapes, and a few common analysis patterns.

```{r}
snapshot <- load_snapshot("Midazolam")
```

## The `as_tibbles()` entry point

`as_tibbles()` takes the snapshot plus a `kind` naming the collection to convert, and returns either a tibble or a named list of tibbles for collections that split into related tables.

Valid values of `kind`:

* `"compounds"`
* `"events"`
* `"expression_profiles"`
* `"formulations"`
* `"individuals"`
* `"observed_data"`
* `"observer_sets"`
* `"populations"`
* `"protocols"`

```{r}
compound_dfs <- as_tibbles(snapshot, "compounds")
names(compound_dfs)
```

The tibble layer is read-only: mutating a returned tibble does not feed changes back into the snapshot.

## Compounds: `$properties` and `$processes`

`as_tibbles(snapshot, "compounds")` returns a list of two tibbles:

* `$properties` carries one row per (compound, parameter) pair across the *Compound* physicochemical properties.
* `$processes` is the long-form view: one row per (compound, process, parameter) triple across every biological *Process* (metabolism, transport, clearance, binding, ...).

```{r}
head(compound_dfs$properties)
head(compound_dfs$processes)
```

The `$processes` tibble carries a `category` column that mirrors `process$category` on the R6 side. Filtering by category gives the same view as the per-category groups on a compound:

```{r, eval=FALSE}
compound_dfs$processes |>
  filter(category == "metabolizing_enzymes")
```

## Individuals: three related tibbles

`as_tibbles(snapshot, "individuals")` returns three tibbles joinable on `individual_id`:

* `$individuals` has one row per *Individual* with its demographic fields (species, population, gender, age, weight, height, gestational age, disease state, calculation methods).
* `$individuals_parameters` is the long-form view of the individual's parameter tree: one row per (individual, parameter path) pair, with value, unit, source, and value-origin metadata.
* `$individuals_expressions` lists the expression profiles attached to each individual: one row per (individual, profile) pair.

```{r}
individual_dfs <- as_tibbles(snapshot, "individuals")
names(individual_dfs)

head(individual_dfs$individuals)
head(individual_dfs$individuals_parameters)
head(individual_dfs$individuals_expressions)
```

## Formulations: `$formulations` and `$formulations_parameters`

`as_tibbles(snapshot, "formulations")` returns two tibbles joinable on `formulation_id`:

* `$formulations` has one row per *Formulation* with its name and release type.
* `$formulations_parameters` is the long-form view of the type-specific release parameters: one row per (formulation, parameter) pair, with value and unit.

```{r}
formulation_dfs <- as_tibbles(snapshot, "formulations")
names(formulation_dfs)
head(formulation_dfs$formulations)
head(formulation_dfs$formulations_parameters)
```

## Populations: `$populations` and `$populations_parameters`

`as_tibbles(snapshot, "populations")` returns two tibbles joinable on `population_id`:

* `$populations` has one row per *Population* with the sampling-recipe header: source population, individual name, sample size, proportion of females, seed, and the age, weight, height, BMI, and eGFR bound columns.
* `$populations_parameters` is the long-form view of the sampling distributions PK-Sim will draw from: one row per (population, parameter) pair, with `distribution_type`, `statistic`, value, and unit.

```{r}
population_dfs <- as_tibbles(snapshot, "populations")
names(population_dfs)
head(population_dfs$populations)
```

## Protocols: `Schema` and `SchemaItem` rows

`as_tibbles(snapshot, "protocols")` returns a single 13-column tibble whether the snapshot has any protocols or not. *Advanced Protocols* contribute one row per *Schema* item, with the protocol-level, schema-level, and item-level parameters joined by name.

```{r}
protocols_df <- as_tibbles(snapshot, "protocols")
head(protocols_df)
```

## Observer sets: `$observer_sets` and `$observers`

`as_tibbles(snapshot, "observer_sets")` returns a list of two tibbles joinable on `observer_set_id` (or `observer_set_name`):

* `$observer_sets` has one row per *ObserverSet* with its name and the count of observers it contains.
* `$observers` has one row per *Observer* with its name, type, dimension, `formula_expression`, `formula_dimension`, `formula_references` (the underlying `ExplicitFormula` references collapsed to `"alias=path"` pairs joined with `|`), and `container_tags` (the `|`-joined `Tag` values from the underlying `ContainerCriteria`).

```{r, eval=FALSE}
observer_dfs <- as_tibbles(snapshot, "observer_sets")
names(observer_dfs)
head(observer_dfs$observers)
```

## Observed data

`as_tibbles(snapshot, "observed_data")` flattens every observed-data series into a long-form tibble ready for plotting.

```{r}
obs_data_df <- as_tibbles(snapshot, "observed_data")
head(obs_data_df)
```

## Next steps

* `vignette("creating-building-blocks")` covers the `create_*()` functions for building new snapshots.
* The full API reference is at <https://esqlabs.github.io/osp.snapshots/reference/>.
