---
title: "Getting started with osp.snapshots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with osp.snapshots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
```

## Overview

osp.snapshots provides an R interface for working with PK-Sim project snapshots. The package lets you:

* Import PK-Sim snapshots from JSON files, URLs, or predefined templates.
* Navigate and explore the nested snapshot structure as R6 objects.
* Modify building blocks and parameters in place.
* Export modified snapshots back to JSON for use in PK-Sim.
* Convert building-block collections to tibbles for analysis in R.

## Loading a snapshot

### From a file, a URL, or a template

`load_snapshot()` dispatches on its input. A local `.json` path is read from disk; an `https?://` URL is fetched; a bare name is resolved against the `OSPSuite.BuildingBlockTemplates` repository.

```{r, eval=FALSE}
# Local file
snapshot <- load_snapshot("path/to/your_snapshot.json")

# Remote URL
snapshot <- load_snapshot("https://example.com/snapshot.json")

# Named template (browse the catalogue with `osp_models()`)
snapshot <- load_snapshot("Midazolam")
```

For this guide we use the `Midazolam` template:

```{r, message=FALSE}
snapshot <- load_snapshot("Midazolam")
snapshot
```

`load_snapshot()` only accepts v11.2 PK-Sim snapshots and newer (`Version >= 79`); older snapshots use pre-v11 conventions that this package does not model. Re-export the project from PK-Sim before loading if needed.

## Exploring building blocks

Building-block collections are stored as named lists on the snapshot. Each `snapshot$<kind>` field is built lazily on first access from the parsed JSON, so the first read of a large collection is the slow one.

### Individuals

An *Individual* is a virtual subject with demographic traits, a physiological parameter tree, and references to expression profiles.

```{r}
# List individuals
snapshot$individuals

# Look up by name
individual <- snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`

# Demographic fields
individual$age
individual$gender
individual$species

# The demographic starting point lives on `$origin_data` (an `OriginData` object)
individual$origin_data
```

### Compounds

A *Compound* exposes its physicochemical properties, its *Calculation methods*, and its biological *Processes*.

```{r}
# List compounds
snapshot$compounds

# Inspect a compound
midazolam <- snapshot$compounds$Midazolam
midazolam

# Calculation methods (read names, add, remove)
midazolam$calculation_methods

# Processes (filter by category to recover the per-category groups)
metab <- Filter(\(p) p$category == "metabolizing_enzymes", midazolam$processes)
names(metab)
```

### Formulations

Formulations describe how a drug is released (`"Dissolved"`, `"Weibull"`, `"First Order"`, `"Zero Order"`, `"Lint80"`, `"Particles"`, `"Table"`).

```{r}
snapshot$formulations
snapshot$formulations$`Tablet (Dormicum)`
```

### Other building blocks

```{r}
# Populations: sampling recipes for cohorts of synthetic subjects
snapshot$populations

# Events: discrete non-administration perturbations
snapshot$events

# Protocols: dosing schedules (simple or advanced with `Schema` blocks)
snapshot$protocols

# Expression profiles: where and how strongly a protein is expressed
snapshot$expression_profiles

# Observer sets: bundles of `Observer` formulas for reuse in simulations
snapshot$observer_sets
```

## Mutating building blocks

Building-block fields are mutable. You can update active bindings directly or use the exported mutator functions; the function form is the public API documented at the package level (the matching R6 methods exist but are an implementation detail).

### Changing fields on a building block

```{r}
# Update age via the active binding
snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`$age <- 45

# Update a parameter value on an `Individual`
ind <- snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`
ind$parameters$`Organism|Gallbladder|Gallbladder ejection fraction`$value <- 0.7

# Verify the change
ind
```

Parameters at a known path inside the individual's parameter tree are exposed as `LocalizedParameter` objects (a subclass of `Parameter`). They carry the full pipe-separated path; v11+ snapshots already write `Events|...`, and the constructor migrates legacy `Applications|...` segments on the way in.

```{r}
params <- snapshot$individuals$`Korean (Yu 2004 study)`$parameters
params$`Organism|Liver|EHC continuous fraction`
params$`Organism|Liver|EHC continuous fraction`$value <- 0.4
```

### Adding and removing building blocks

Each building-block kind has a pair of exported mutators that take the snapshot first and return it invisibly so they chain with the base pipe:

```{r, eval=FALSE}
# Mutator functions chain
snapshot |>
  add_individual(create_individual("Patient_A", age = 35, gender = "FEMALE")) |>
  remove_population("Some population")
```

See `vignette("creating-building-blocks")` for factory recipes that feed these mutators.

## Working with observed data

Observed data live on `snapshot$observed_data`. Each entry is an `ospsuite::DataSet`, ready to feed the ospsuite analysis pipeline.

```{r}
snapshot$observed_data
obs_data <- snapshot$observed_data[[1]]
obs_data
class(obs_data)
```

`export_snapshot()` replays the observed-data section from the original JSON (filtered to entries still present after `remove_observed_data()`), so mutations on a `DataSet` after load are not preserved on export.

## Converting collections to tibbles

`as_tibbles()` is the canonical entry point for the tibble layer. It takes the snapshot plus a `kind` and returns either a bare tibble (for `"protocols"` and `"observed_data"`) or a named list of related tibbles (for every other kind).

```{r}
# Bare-tibble collections (e.g. protocols)
head(as_tibbles(snapshot, "protocols"))

# Multi-tibble collections (compounds, individuals, formulations, populations, observer sets, ...)
compound_dfs <- as_tibbles(snapshot, "compounds")
names(compound_dfs)
```

The eight `get_*_dfs()` functions stay available as thin wrappers around `as_tibbles()`. See `vignette("working-with-dataframes")` for the full tibble layer.

## Exporting

After mutating a snapshot, write it back to JSON with `export_snapshot()`:

```{r, eval=FALSE}
export_snapshot(snapshot, "modified_snapshot.json")
```

The exported file imports back into PK-Sim. Top-level JSON sections that osp.snapshots does not model (e.g. `Simulations`, `Classifications`, `ParameterIdentifications`) pass through unchanged.

## Next steps

* `vignette("creating-building-blocks")` walks through the `create_*()` factory family for building new snapshots from scratch.
* `vignette("working-with-dataframes")` covers the tibble layer in depth.
