---
title: "Getting started with osp.snapshots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with osp.snapshots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
```

## Overview

osp.snapshots provides an R interface for working with PK-Sim project snapshots. The package lets you:

* Import PK-Sim snapshots from JSON files, URLs, or predefined templates.
* Navigate and explore the nested snapshot structure as R6 objects.
* Modify building blocks and parameters in place.
* Export modified snapshots back to JSON for use in PK-Sim.
* Convert building-block collections to tibbles for analysis in R.

## Loading a snapshot

### From a file, a URL, or a template

`load_snapshot()` dispatches on its input. A local `.json` path is read from disk; an `https?://` URL is fetched; a bare name is resolved against the `OSPSuite.BuildingBlockTemplates` repository.

```{r, eval=FALSE}
# Local file
snapshot <- load_snapshot("path/to/your_snapshot.json")

# Remote URL
snapshot <- load_snapshot("https://example.com/snapshot.json")

# Named template (browse the catalogue with `osp_models()`)
snapshot <- load_snapshot("Midazolam")
```

For this guide we use the `Midazolam` template:

```{r, message=FALSE}
snapshot <- load_snapshot("Midazolam")
snapshot
```

`load_snapshot()` only accepts v11.2 PK-Sim snapshots and newer (`Version >= 79`); older snapshots use pre-v11 conventions that this package does not model. Re-export the project from PK-Sim before loading if needed.

## Exploring building blocks

A loaded snapshot is a navigable tree: each kind of building block (individuals, compounds, formulations, populations, events, protocols, expression profiles, observer sets) lives in a named list under `snapshot$<kind>` that you can list, look up by name, and inspect interactively. Printing any object prints a readable summary; tab-completion works the whole way down.

### Walking one kind: individuals

`snapshot$individuals` is a named list. Print it for a short overview, ask for `names()` and `length()` to see what is there, then index by name to get a single building block:

```{r}
snapshot$individuals

names(snapshot$individuals)
length(snapshot$individuals)

individual <- snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`
individual
```

Once you have an individual, the fields you would type in PK-Sim are available as active bindings on the object:

```{r}
individual$age
individual$age_unit
individual$gender
individual$species
individual$weight
```

Nested structures are objects themselves, so you can keep drilling. The demographic starting point lives on `$origin_data`, and the calculation methods PK-Sim will use to create the virtual individual sit one level deeper:

```{r}
individual$origin_data

individual$origin_data$calculation_methods
individual$origin_data$calculation_methods$names
```

Path-bearing parameters on the individual are exposed as a named list keyed by their PK-Sim path, so you can list them and pull one out by path:

```{r}
korean <- snapshot$individuals$`Korean (Yu 2004 study)`
names(korean$parameters)
korean$parameters$`Organism|Liver|EHC continuous fraction`
```

### The same shape for every other kind

Every other building-block collection works the same way: a named list under `snapshot$<kind>` that you can list, index by name, and print. The objects you get back have their own fields and nested structures, all available through `$`. For example, compounds expose their physicochemical properties, their calculation methods, and their biological processes:

```{r}
names(snapshot$compounds)
midazolam <- snapshot$compounds$Midazolam
midazolam$molecular_weight
midazolam$is_small_molecule
midazolam$calculation_methods$names
names(midazolam$processes)
```

Other collections feel just as familiar. Asking for `names()` is usually the first move:

```{r}
names(snapshot$formulations)
length(snapshot$protocols)
head(names(snapshot$protocols))
names(snapshot$expression_profiles)
```

And every kind of building block can be printed for a readable summary:

```{r}
snapshot$formulations$`Tablet (Dormicum)`
snapshot$protocols$`po 15 mg`
```

## Mutating building blocks

Most fields on a building block are mutable, and every kind has a pair of `add_*()` / `remove_*()` functions for inserting or pruning entries. These functions are the public API; the matching R6 methods exist but are an implementation detail.

### Changing fields on a building block

Assigning to a field on an existing object writes the change straight back into the snapshot:

```{r}
ind <- snapshot$individuals$`European (P-gp modified, CYP3A4 36 h)`
ind$age <- 45
ind$weight <- 78
ind
```

Parameters that sit at a known path inside the individual's parameter tree are exposed as `LocalizedParameter` objects with a pipe-separated path. v11+ snapshots already write `Events|...` segments; legacy `Applications|...` segments are migrated on load.

```{r}
korean <- snapshot$individuals$`Korean (Yu 2004 study)`
ehc <- korean$parameters$`Organism|Liver|EHC continuous fraction`
ehc$value <- 0.4
ehc
```

### Adding and removing building blocks

Build the new object with the matching `create_*()` function (see `vignette("creating-building-blocks")`), then attach it with `add_*()`. Removal is by name.

```{r}
patient <- create_individual("Patient_A", age = 35, gender = "FEMALE")
add_individual(snapshot, patient)

remove_individual(snapshot, "Patient_A")
```

### Chaining with the base pipe

`add_*()` and `remove_*()` take the snapshot first and return it invisibly, so you can chain them with the base pipe when you want to make several changes in one expression:

```{r, eval=FALSE}
snapshot |>
  add_individual(create_individual("Patient_A", age = 35, gender = "FEMALE")) |>
  add_individual(create_individual("Patient_B", age = 60, gender = "MALE")) |>
  remove_population("Healthy Adults")
```

## Converting collections to tibbles

`as_tibbles()` converts any building-block collection to a tibble (or a named list of related tibbles for collections that split into more than one). It takes the snapshot plus the `kind` to convert:

```{r}
head(as_tibbles(snapshot, "protocols"))

compound_dfs <- as_tibbles(snapshot, "compounds")
names(compound_dfs)
```

See `vignette("exporting-snapshots-to-dataframes")` for the per-kind shapes and analysis examples.

## Exporting

After mutating a snapshot, write it back to JSON with `export_snapshot()`:

```{r, eval=FALSE}
export_snapshot(snapshot, "modified_snapshot.json")
```

The exported file imports back into PK-Sim. Top-level JSON sections that osp.snapshots does not model (e.g. `Simulations`, `Classifications`, `ParameterIdentifications`) pass through unchanged.

## Next steps

* `vignette("creating-building-blocks")` walks through the `create_*()` functions for building new snapshots from scratch.
* `vignette("exporting-snapshots-to-dataframes")` covers the tibble layer in depth.
