---
title: "Building simulations from scratch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building simulations from scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
```

## Overview

This vignette shows how to assemble a *Simulation* from named arguments and attach it to a *Snapshot*. A simulation binds together other building blocks (an individual or a population, compounds, events, observer sets, observed data, protocols, formulations) plus a solver, an output schema, and parameter overrides.

We start with the `Midazolam` template, which already ships individuals, compounds, protocols, formulations, and events that we can reference by name.

```{r}
snapshot <- load_snapshot("Midazolam")
names(snapshot$compounds)
names(snapshot$individuals)
names(snapshot$protocols)
names(snapshot$formulations)
names(snapshot$events)
```

## `create_simulation()` arguments at a glance

Every argument falls into one of three kinds: a plain scalar, a character vector of names that reference existing building blocks, or a structured object built by a matching `create_*()` sub-factory. The sections that follow walk through each sub-factory; this table is the map.

| Argument              | Type                                                       | Required? |
| --------------------- | ---------------------------------------------------------- | --------- |
| `name`                | string                                                     | yes       |
| `individual`          | string (name of an Individual building block)              | XOR with `population` |
| `population`          | string (name of a Population building block)               | XOR with `individual` |
| `model`               | string (defaults to `"4Comp"`)                             | no        |
| `description`         | string                                                     | no        |
| `allow_aging`         | logical                                                    | no        |
| `observed_data_names` | character vector of observed-data names                    | no        |
| `output_selections`   | character vector of output-quantity paths                  | no        |
| `solver`              | `create_solver_settings()` output                          | no        |
| `output_schema`       | `create_output_schema()` output                            | no        |
| `compounds`           | list of `create_compound_properties()` outputs             | no        |
| `events`              | list of `create_event_selection()` outputs                 | no        |
| `observer_sets`       | list of `create_observer_set_selection()` outputs          | no        |
| `output_mappings`     | list of `create_output_mapping()` outputs                  | no        |
| `parameters`          | list of `create_parameter(path = ...)` outputs             | no        |
| `advanced_parameters` | list of `AdvancedParameter` objects (population sims only) | no        |

**Exactly one** of `individual` or `population` must be supplied (passing both, or neither, errors).

### What PK-Sim does when an optional argument is omitted

- `model`: defaults to `"4Comp"` (the standard four-compartment model).
- `description`: stays empty; no functional effect on the simulation.
- `allow_aging`: defaults to `FALSE` on load (no ageing during the simulation window).
- `observed_data_names`: no observed-data sets are attached to the simulation.
- `output_selections`: no quantities are explicitly marked for reporting beyond whatever the model defaults to. Supply at least one path to control what the simulation outputs.
- `solver`: PK-Sim builds a `SolverSettings` from `ISolverSettingsFactory.CreateDefault()` (built-in CVODE defaults).
- `output_schema`: PK-Sim builds an empty schema via `IOutputSchemaFactory.CreateEmpty()`; the simulation reports no time points. Supply at least one interval to get output.
- `compounds`: the simulation has no compound configuration. Compounds referenced indirectly (e.g. via a protocol) still need a `CompoundProperties` entry to be runnable.
- `events`: no events are attached.
- `observer_sets`: no observer sets are attached.
- `output_mappings`: no observed-data-to-output bindings are recorded. Parameter identification and downstream comparison tools will have nothing to compare.
- `parameters`: no parameter overrides are applied; every quantity uses the value derived from its building block.
- `advanced_parameters`: no variability overrides on population sampling; default distributions apply.

## Solver settings

`create_solver_settings()` produces a sparse solver configuration.

**All arguments are optional**, with the PK-Sim fallback when omitted:

- `abs_tol`, `rel_tol`, `use_jacobian`, `h0`, `h_min`, `h_max`, `mx_step`: each field omitted from the snapshot is filled by `ISolverSettingsFactory.CreateDefault()` on load, so the simulation runs with PK-Sim's built-in CVODE defaults. Calling `create_solver_settings()` with no arguments is the same as not supplying `solver` to `create_simulation()` at all.

```{r}
solver <- create_solver_settings(
  abs_tol = 1e-9,
  rel_tol = 1e-9,
  mx_step = 100000
)
```

## Output schema

`create_output_schema()` collects one or more *Output intervals*.

- **Required:** none.
- **Optional**, with the PK-Sim fallback when omitted:
  - `intervals`: defaults to an empty list. An output schema with no intervals means the simulation reports no time points; supply at least one interval if you want any output.

For `create_output_interval()`:

- **Required:** `start_time`, `end_time`, `resolution`, all numeric scalars in PK-Sim's default units (`h`, `h`, `pts/h`).
- **Optional**, with the PK-Sim fallback when omitted:
  - `name`: PK-Sim assigns a generated unique name through `IContainerTask.CreateUniqueName()`.

```{r}
schema <- create_output_schema(
  intervals = list(
    create_output_interval(start_time = 0, end_time = 2, resolution = 20),
    create_output_interval(start_time = 2, end_time = 24, resolution = 20)
  )
)
```

## Compound properties

`create_compound_properties()` configures one compound for the simulation.

### Compound vs CompoundProperties

A `Compound` building block (the kind you find in `snapshot$compounds`) is the project-level definition of a drug: its physicochemical properties, the full menu of available alternatives, the process templates it ships, and so on. The same compound can be referenced by many simulations.

`CompoundProperties` is the *per-simulation overlay* that picks one path through that menu. For each simulation that uses the compound you record:

- which `calculation_methods` override the compound's defaults,
- which `alternatives` (e.g. solubility, permeability) are selected,
- which `processes` are enabled, and how each is wired to molecules or systemic types,
- which `protocol` is used and which formulation fills each application slot.

None of those choices live on the `Compound` building block, which is why `create_compound_properties()` exists as a separate step even when the compound is already in `snapshot$compounds`. The reference back to the building block is the `name` argument; everything else is simulation-local.

By contrast, `create_event_selection()` only takes `name` plus a `start_time`, and `create_observer_set_selection()` only takes `name`, because events and observer sets have nothing analogous to alternatives/processes/protocol to choose between at simulation time.

### Arguments

- **Required:** `name`, which must match a compound in the snapshot.
- **Optional**, with the PK-Sim fallback when omitted:
  - `calculation_methods`: the simulation uses the calculation-method cache already defined on the `Compound` building block.
  - `alternatives`: each unspecified alternative group keeps the default alternative marked on the compound (typically the first or the one with `IsDefault = TRUE`).
  - `processes`: every compound process that is not listed is treated as deselected; PK-Sim materialises the right placeholder (`NotSelectedSystemicProcess`, `NoInteractionProcess`, ...) so the simulation runs as if you had explicitly disabled it.
  - `protocol`: when omitted, the compound is included with no dosing protocol attached, which means it will not be administered during the simulation (useful for compounds that are only produced metabolically).

```{r}
compound <- create_compound_properties(
  name = "Midazolam",
  calculation_methods = c(
    "Cellular partition coefficient method - Rodgers and Rowland",
    "Cellular permeability - PK-Sim Standard"
  ),
  processes = list(
    create_compound_process_selection(systemic_process_type = "Hepatic")
  ),
  protocol = create_protocol_selection(
    name = "iv 1 mg (bolus)",
    formulations = list(
      create_formulation_selection(
        name = "Oral solution",
        key = "Formulation"
      )
    )
  )
)
```

The supporting factories follow the same convention:

- `create_compound_process_selection()`: every field is optional; PK-Sim resolves the right placeholder from whichever combination of `name`, `molecule_name`, `metabolite_name`, `compound_name`, and `systemic_process_type` is supplied. Does not reference a building block; the `name` field here is a process name internal to the parent compound.
- `create_compound_group_selection()`: `group_name` and `alternative_name` are both required. Does not reference a building block; it picks an alternative inside an alternative group defined on the parent compound.

### Protocol selection: Protocol vs ProtocolSelection

`create_protocol_selection()` mirrors the same split as compounds. A `Protocol` building block (in `snapshot$protocols`) is the project-level dosing schedule with all its parameters. `ProtocolSelection` is the per-simulation overlay that points at one such protocol by `name` and records which `Formulation` fills each of the protocol's application slots through `formulations` (a list of `create_formulation_selection()` outputs).

- **Required:** `name` (must match a protocol in the snapshot).
- **Optional**, with the PK-Sim fallback when omitted:
  - `formulations`: protocol application slots that have no formulation binding stay empty; PK-Sim will reject the simulation at load time if the protocol requires a formulation for an unbound slot. Supply `formulations` for every slot that needs one (typically every slot in oral protocols).

### Formulation selection: Formulation vs FormulationSelection

Same idea again. A `Formulation` building block (in `snapshot$formulations`) is the project-level dissolution profile. `FormulationSelection` is the per-simulation entry that binds that formulation to one application slot inside the chosen protocol.

- **Required:** `name` (must match a formulation in the snapshot) and `key` (the application slot identifier inside the protocol).

## Event and observer-set selections

### Event selection: Event vs EventSelection

An `Event` building block (in `snapshot$events`) is the project-level event definition (template plus parameters). `EventSelection` is the per-simulation reference that picks one event by `name` and supplies its `start_time` for this simulation. Aside from `start_time`, events carry no per-simulation overlay.

`create_event_selection()`:

- **Required:** `name` (must match an event building block) and `start_time` (numeric, hours).

### Observer-set selection: ObserverSet vs ObserverSetSelection

An `ObserverSet` building block (in `snapshot$observer_sets`) is the project-level observer bundle. `ObserverSetSelection` is the per-simulation reference that picks one observer set by `name`. There is no per-simulation overlay at all, which is why the factory takes only a single argument.

`create_observer_set_selection()`:

- **Required:** `name` (must match an observer-set building block).

```{r}
event <- create_event_selection(name = "High-fat breakfast", start_time = 12)
```

The `Midazolam` template ships no observer sets, so we skip `create_observer_set_selection()` in this walkthrough. With a template that does have one (`snapshot$observer_sets`), the call would look like:

```{r, eval = FALSE}
observer_set <- create_observer_set_selection(
  name = "BrainPlasmaConcentration"
)
```

## Optional parameter overrides

Simulation parameter overrides are *localized parameters*, identified by their full container path inside the simulation's parameter tree. Use `create_parameter(path = ...)` to obtain one.

```{r}
overrides <- list(
  create_parameter(
    path = "Organism|Liver|Volume",
    value = 1.8,
    unit = "L"
  )
)
```

## Assemble the simulation

With every input prepared, the final `create_simulation()` call wires them together.

```{r}
sim <- create_simulation(
  name = "Midazolam single dose",
  model = "4Comp",
  individual = "Korean (Yu 2004 study)",
  compounds = list(compound),
  events = list(event),
  solver = solver,
  output_schema = schema,
  output_selections = c(
    "Organism|PeripheralVenousBlood|Midazolam|Plasma (Peripheral Venous Blood)"
  ),
  parameters = overrides
)

sim
```

## Attach and export

`add_simulation()` appends the simulation to a snapshot. If any name referenced by the simulation does not resolve to a known building block, you get one informational warning per simulation; the add proceeds either way.

```{r}
snapshot <- snapshot |>
  add_simulation(sim)
```

Export with `export_snapshot()` and PK-Sim will load the resulting JSON.

```{r, eval = FALSE}
export_snapshot(snapshot, "midazolam-with-new-sim.json")
```

## Removing a simulation

```{r}
snapshot <- snapshot |>
  remove_simulation("Midazolam single dose")
```
