---
title: "Building simulations from scratch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building simulations from scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
```

## Overview

This vignette shows how to build a *Simulation* and attach it to a *Snapshot* in a single `add_simulation()` call. A simulation binds together other building blocks (an individual or a population, compounds, events, observer sets, observed data, protocols, formulations) plus a solver, an output schema, and parameter overrides. Because a simulation is almost pure references into the snapshot, the snapshot is its entry point: with every building block already in hand, `add_simulation()` resolves those references and derives sensible defaults for you.

We start with the `Midazolam` template, which already ships individuals, compounds, protocols, formulations, and events that we can reference by name.

```{r}
snapshot <- load_snapshot("Midazolam")
names(snapshot$compounds)
names(snapshot$individuals)
names(snapshot$protocols)
names(snapshot$formulations)
names(snapshot$events)
```

## `add_simulation()` arguments at a glance

`add_simulation()` takes the `snapshot` as its first argument, then a `name` and the building-block references that make up the simulation. Every argument falls into one of three kinds: a plain scalar, a character vector of names that reference existing building blocks, or a structured value (most often an inline `compounds` config; see below). The sections that follow walk through each piece; this table is the map.

| Argument              | Type                                                       | Required? |
| --------------------- | ---------------------------------------------------------- | --------- |
| `snapshot`            | a `Snapshot` (the entry point; first argument)             | yes       |
| `name`                | string                                                     | yes       |
| `individual`          | string (name of an Individual building block)              | XOR with `population` |
| `population`          | string (name of a Population building block)               | XOR with `individual` |
| `model`               | string (defaults to `"4Comp"`)                             | no        |
| `description`         | string                                                     | no        |
| `allow_aging`         | logical                                                    | no        |
| `observed_data_names` | character vector of observed-data names                    | no        |
| `output_selections`   | character vector of output-quantity paths                  | no        |
| `solver`              | `create_solver_settings()` output                          | no        |
| `output_schema`       | `create_output_schema()` output                            | no        |
| `compounds`           | list of inline compound-config lists (`list(name =, protocol =, formulation =, processes =, ...)`) and/or `create_compound_properties()` outputs | no |
| `events`              | list of `create_event_selection()` outputs                 | no        |
| `observer_sets`       | list of `create_observer_set_selection()` outputs          | no        |
| `output_mappings`     | list of `create_output_mapping()` outputs                  | no        |
| `parameters`          | list of `create_parameter(path = ...)` outputs             | no        |
| `advanced_parameters` | list of `AdvancedParameter` objects (population sims only) | no        |

**Exactly one** of `individual` or `population` must be supplied (passing both, or neither, errors).

### What PK-Sim does when an optional argument is omitted

- `model`: defaults to `"4Comp"` (the standard four-compartment model).
- `description`: stays empty; no functional effect on the simulation.
- `allow_aging`: defaults to `FALSE` on load (no ageing during the simulation window).
- `observed_data_names`: no observed-data sets are attached to the simulation.
- `output_selections`: no quantities are explicitly marked for reporting beyond whatever the model defaults to. Supply at least one path to control what the simulation outputs.
- `solver`: PK-Sim builds a `SolverSettings` from `ISolverSettingsFactory.CreateDefault()` (built-in CVODE defaults).
- `output_schema`: PK-Sim builds an empty schema via `IOutputSchemaFactory.CreateEmpty()`; the simulation reports no time points. Supply at least one interval to get output.
- `compounds`: the simulation has no compound configuration. Compounds referenced indirectly (e.g. via a protocol) still need a compound entry to be runnable.
- `events`: no events are attached.
- `observer_sets`: no observer sets are attached.
- `output_mappings`: no observed-data-to-output bindings are recorded. Parameter identification and downstream comparison tools will have nothing to compare.
- `parameters`: no parameter overrides are applied; every quantity uses the value derived from its building block.
- `advanced_parameters`: no variability overrides on population sampling; default distributions apply.

## Solver settings

`create_solver_settings()` produces a sparse solver configuration.

**All arguments are optional**, with the PK-Sim fallback when omitted:

- `abs_tol`, `rel_tol`, `use_jacobian`, `h0`, `h_min`, `h_max`, `mx_step`: each field omitted from the snapshot is filled by `ISolverSettingsFactory.CreateDefault()` on load, so the simulation runs with PK-Sim's built-in CVODE defaults. Calling `create_solver_settings()` with no arguments is the same as not supplying `solver` to `add_simulation()` at all.

```{r}
solver <- create_solver_settings(
  abs_tol = 1e-9,
  rel_tol = 1e-9,
  mx_step = 100000
)
```

## Output schema

`create_output_schema()` collects one or more *Output intervals*.

- **Required:** none.
- **Optional**, with the PK-Sim fallback when omitted:
  - `intervals`: defaults to an empty list. An output schema with no intervals means the simulation reports no time points; supply at least one interval if you want any output.

For `create_output_interval()`:

- **Required:** `start_time`, `end_time`, `resolution`, all numeric scalars in PK-Sim's default units (`h`, `h`, `pts/h`).
- **Optional**, with the PK-Sim fallback when omitted:
  - `name`: PK-Sim assigns a generated unique name through `IContainerTask.CreateUniqueName()`.

```{r}
schema <- create_output_schema(
  intervals = list(
    create_output_interval(start_time = 0, end_time = 2, resolution = 20),
    create_output_interval(start_time = 2, end_time = 24, resolution = 20)
  )
)
```

## Compound properties

Each entry of `compounds` configures one compound for the simulation. The common path is an inline config list; the escape-hatch factories are still accepted through the same slot (see below).

### Compound vs compound configuration

A `Compound` building block (the kind you find in `snapshot$compounds`) is the project-level definition of a drug: its physicochemical properties, the full menu of available alternatives, the process templates it ships, and so on. The same compound can be referenced by many simulations.

A compound *configuration* is the per-simulation overlay that picks one path through that menu. For each simulation that uses the compound you record:

- which `calculation_methods` override the compound's defaults,
- which `alternatives` (e.g. solubility, permeability) are selected,
- which `processes` are enabled, and how each is wired to molecules or systemic types,
- which `protocol` is used and which formulation fills each application slot.

Because the snapshot is the entry point, `add_simulation()` derives most of these from the referenced building block when you leave them out: you point at the compound by `name`, and the resolver fills in the rest.

### Inline fields

- **Required:** `name`, which must match a compound in the snapshot.
- **Optional**, derived from the snapshot when omitted:
  - `calculation_methods`: derived from the calculation methods defined on the referenced `Compound` building block.
  - `alternatives`: each alternative group is defaulted to that group's default alternative (the one marked `IsDefault`, or `"User defined"` as a fallback).
  - `formulation` (together with `protocol`): the resolver infers the formulation key from the referenced protocol's application slot (falling back to the literal `"Formulation"` key for simple protocols).
  - `processes`: **never defaulted.** Omitting `processes` produces no process array, which deselects the compound's processes in PK-Sim (it materialises the right placeholder so the simulation runs as if you had explicitly disabled them). List the processes you want to keep. A name is classified as a systemic-process type when it is one of `"Hepatic"`, `"Renal"`, or `"Biliary"`; every other name is treated as a molecule name.
  - `protocol`: when omitted, the compound is included with no dosing protocol attached, which means it will not be administered during the simulation (useful for compounds that are only produced metabolically).

```{r}
compound <- list(
  name = "Midazolam",
  protocol = "iv 1 mg (bolus)",
  formulation = "Oral solution",
  processes = c("Hepatic")
)
```

### The escape hatch

For a multi-slot protocol, or any hand-built configuration you want full control over, build the compound entry with the factory functions and pass the resulting `CompoundProperties` object through the same `compounds` slot. When you do, the resolver leaves it untouched (no defaulting):

```{r, eval = FALSE}
compound <- create_compound_properties(
  name = "Midazolam",
  calculation_methods = c(
    "Cellular partition coefficient method - Rodgers and Rowland",
    "Cellular permeability - PK-Sim Standard"
  ),
  processes = list(
    create_compound_process_selection(systemic_process_type = "Hepatic")
  ),
  protocol = create_protocol_selection(
    name = "iv 1 mg (bolus)",
    formulations = list(
      create_formulation_selection(
        name = "Oral solution",
        key = "Formulation"
      )
    )
  )
)
```

The supporting factories follow the same convention:

- `create_compound_process_selection()`: every field is optional; PK-Sim resolves the right placeholder from whichever combination of `name`, `molecule_name`, `metabolite_name`, `compound_name`, and `systemic_process_type` is supplied. Does not reference a building block; the `name` field here is a process name internal to the parent compound.
- `create_compound_group_selection()`: `group_name` and `alternative_name` are both required. Does not reference a building block; it picks an alternative inside an alternative group defined on the parent compound.

### Protocol selection: Protocol vs ProtocolSelection

`create_protocol_selection()` mirrors the same split as compounds. A `Protocol` building block (in `snapshot$protocols`) is the project-level dosing schedule with all its parameters. `ProtocolSelection` is the per-simulation overlay that points at one such protocol by `name` and records which `Formulation` fills each of the protocol's application slots through `formulations` (a list of `create_formulation_selection()` outputs).

- **Required:** `name` (must match a protocol in the snapshot).
- **Optional**, with the PK-Sim fallback when omitted:
  - `formulations`: protocol application slots that have no formulation binding stay empty; PK-Sim will reject the simulation at load time if the protocol requires a formulation for an unbound slot. Supply `formulations` for every slot that needs one (typically every slot in oral protocols).

### Formulation selection: Formulation vs FormulationSelection

Same idea again. A `Formulation` building block (in `snapshot$formulations`) is the project-level dissolution profile. `FormulationSelection` is the per-simulation entry that binds that formulation to one application slot inside the chosen protocol.

- **Required:** `name` (must match a formulation in the snapshot) and `key` (the application slot identifier inside the protocol).

## Event and observer-set selections

### Event selection: Event vs EventSelection

An `Event` building block (in `snapshot$events`) is the project-level event definition (template plus parameters). `EventSelection` is the per-simulation reference that picks one event by `name` and supplies its `start_time` for this simulation. Aside from `start_time`, events carry no per-simulation overlay.

`create_event_selection()`:

- **Required:** `name` (must match an event building block) and `start_time` (numeric, hours).

### Observer-set selection: ObserverSet vs ObserverSetSelection

An `ObserverSet` building block (in `snapshot$observer_sets`) is the project-level observer bundle. `ObserverSetSelection` is the per-simulation reference that picks one observer set by `name`. There is no per-simulation overlay at all, which is why the factory takes only a single argument.

`create_observer_set_selection()`:

- **Required:** `name` (must match an observer-set building block).

```{r}
event <- create_event_selection(name = "High-fat breakfast", start_time = 12)
```

The `Midazolam` template ships no observer sets, so we skip `create_observer_set_selection()` in this walkthrough. With a template that does have one (`snapshot$observer_sets`), the call would look like:

```{r, eval = FALSE}
observer_set <- create_observer_set_selection(
  name = "BrainPlasmaConcentration"
)
```

## Optional parameter overrides

Simulation parameter overrides are *localized parameters*, identified by their full container path inside the simulation's parameter tree. Use `create_parameter(path = ...)` to obtain one.

```{r}
overrides <- list(
  create_parameter(
    path = "Organism|Liver|Volume",
    value = 1.8,
    unit = "L"
  )
)
```

## Build, attach, and export

With every input prepared, a single `add_simulation()` call builds the simulation from these references, resolves them against the snapshot, derives the compound defaults, and attaches the result. If any name referenced by the simulation does not resolve to a known building block, you get one informational warning per simulation; the add proceeds either way.

```{r}
snapshot <- snapshot |>
  add_simulation(
    name = "Midazolam single dose",
    model = "4Comp",
    individual = "Korean (Yu 2004 study)",
    compounds = list(compound),
    events = list(event),
    solver = solver,
    output_schema = schema,
    output_selections = c(
      "Organism|PeripheralVenousBlood|Midazolam|Plasma (Peripheral Venous Blood)"
    ),
    parameters = overrides
  )

snapshot$simulations[["Midazolam single dose"]]
```

Export with `export_snapshot()` and PK-Sim will load the resulting JSON.

```{r, eval = FALSE}
export_snapshot(snapshot, "midazolam-with-new-sim.json")
```

## Removing a simulation

```{r}
snapshot <- snapshot |>
  remove_simulation("Midazolam single dose")
```
