---
title: "Working with data frames"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with data frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
library(dplyr)
```

## Overview

osp.snapshots ships a tibble layer that converts the nested PK-Sim snapshot tree into flat, joinable tibbles ready for downstream analysis with `dplyr` and friends. This vignette walks through the entry point, the per-collection shapes, and a few common analysis patterns.

```{r}
snapshot <- load_snapshot("Midazolam")
```

## The `as_tibbles()` entry point

`as_tibbles()` is the canonical entry point. It takes the snapshot plus a `kind` naming the collection to convert, and returns either a tibble or a named list of tibbles for collections that split into related tables.

Valid values of `kind`:

* `"compounds"`
* `"events"`
* `"expression_profiles"`
* `"formulations"`
* `"individuals"`
* `"observed_data"`
* `"observer_sets"`
* `"populations"`
* `"protocols"`

```{r}
compound_dfs <- as_tibbles(snapshot, "compounds")
names(compound_dfs)
```

The eight `get_*_dfs()` helpers (`get_compounds_dfs()`, `get_individuals_dfs()`, `get_formulations_dfs()`, `get_populations_dfs()`, `get_events_dfs()`, `get_expression_profiles_dfs()`, `get_protocols_dfs()`, `get_observed_data_dfs()`) plus `get_observer_sets_dfs()` remain available as thin wrappers around `as_tibbles()`. New code should prefer `as_tibbles()`.

The tibble layer is read-only: mutating a returned tibble does not feed changes back into the snapshot.

## Compounds: `$properties` and `$processes`

`get_compounds_dfs()` (and `as_tibbles(snapshot, "compounds")`) returns a list of two tibbles:

* `$properties` carries one row per (compound, parameter) pair across the *Compound* physicochemical properties.
* `$processes` is the long-form view: one row per (compound, process, parameter) triple across every biological *Process* (metabolism, transport, clearance, binding, ...).

```{r}
head(compound_dfs$properties)
head(compound_dfs$processes)
```

The `$processes` tibble carries a `category` column that mirrors `process$category` on the R6 side. Filtering by category gives the same view as the deprecated per-category accessors on `Compound`:

```{r, eval=FALSE}
compound_dfs$processes |>
  filter(category == "metabolizing_enzymes")
```

## Individuals: three related tibbles

`as_tibbles(snapshot, "individuals")` returns three tibbles joinable on `individual_id`:

```{r}
individual_dfs <- as_tibbles(snapshot, "individuals")
names(individual_dfs)

head(individual_dfs$individuals)
head(individual_dfs$individuals_parameters)
head(individual_dfs$individuals_expressions)
```

You can also convert a single `Individual` directly via its `$to_df()` method:

```{r}
individual <- snapshot$individuals[[1]]
single_dfs <- individual$to_df()
names(single_dfs)
head(single_dfs$individuals)
```

## Formulations

```{r}
formulation_dfs <- as_tibbles(snapshot, "formulations")
names(formulation_dfs)
head(formulation_dfs$formulations)
head(formulation_dfs$formulations_parameters)
```

## Populations

```{r}
population_dfs <- as_tibbles(snapshot, "populations")
names(population_dfs)
head(population_dfs$populations)
```

## Protocols: `Schema` and `SchemaItem` rows

`as_tibbles(snapshot, "protocols")` returns a 13-column tibble whether the snapshot has any protocols or not (this used to vary; see NEWS for #56). *Advanced Protocols* contribute one row per *Schema* item, with the protocol-level, schema-level, and item-level parameters joined by name.

```{r}
protocols_df <- as_tibbles(snapshot, "protocols")
head(protocols_df)
```

## Observer sets: `$observer_sets` and `$observers`

`as_tibbles(snapshot, "observer_sets")` returns a list of two tibbles:

* `$observer_sets` has one row per *ObserverSet* with its name and observer count.
* `$observers` has one row per *Observer* with `observer_set_id` / `observer_set_name` columns for joining back to the parent set.

```{r, eval=FALSE}
observer_dfs <- as_tibbles(snapshot, "observer_sets")
names(observer_dfs)
head(observer_dfs$observers)
```

The `container_tags` column on the `$observers` tibble carries the `|`-joined `Tag` values from the underlying `ContainerCriteria`.

## Observed data

`as_tibbles(snapshot, "observed_data")` flattens every observed-data series into a long-form tibble ready for plotting.

```{r}
obs_data_df <- as_tibbles(snapshot, "observed_data")
head(obs_data_df)
```

## Next steps

* `vignette("creating-building-blocks")` covers the `create_*()` factory family for building new snapshots.
* The full API reference is at <https://esqlabs.github.io/osp.snapshots/reference/>.
