---
title: "Working with Data Frames"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Data Frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(osp.snapshots)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Overview

One of the key features of osp.snapshots is the ability to convert complex nested snapshot data into structured data frames that are easy to work with in R. This vignette demonstrates how to extract, convert, and analyze snapshot data using standard R data analysis tools.

## Setup

Let's start by loading our test snapshot:

```{r}
snapshot <- load_snapshot("Midazolam")
```

## Converting Building Blocks to Data Frames

The unified `as_tibbles()` function is the canonical entry point. It takes a snapshot plus a `kind` naming the collection to convert, and returns a tibble (or named list of tibbles for collections that need to be split into related tables).

Valid values of `kind` are `"compounds"`, `"individuals"`, `"formulations"`, `"populations"`, `"events"`, `"expression_profiles"`, `"protocols"`, and `"observed_data"`.

### Converting Individuals

Individual data is converted into multiple related data frames:

```{r}
individual_dfs <- as_tibbles(snapshot, "individuals")
names(individual_dfs)
```

Let's examine each data frame:

```{r}
# Basic characteristics
head(individual_dfs$individuals)

# Parameters
head(individual_dfs$individuals_parameters)

# Expression profiles
head(individual_dfs$individuals_expressions)
```

### Converting Single Individual

You can also convert a single individual to data frames:

```{r}
# Get the first individual
individual <- snapshot$individuals[[1]]
single_individual_df <- individual$to_df()

names(single_individual_df)
head(single_individual_df$individuals)
```

### Converting Compounds

```{r}
compounds_df <- as_tibbles(snapshot, "compounds")
head(compounds_df)
```

### Converting Formulations

```{r}
formulations_dfs <- as_tibbles(snapshot, "formulations")
names(formulations_dfs)

# Basic formulation info
head(formulations_dfs$formulations)

# Formulation parameters
head(formulations_dfs$formulations_parameters)
```

### Converting Populations

```{r}
populations_dfs <- as_tibbles(snapshot, "populations")
names(populations_dfs)

head(populations_dfs$populations)
```

### Converting Observed Data

Observed data conversion creates a flat data frame perfect for analysis:

```{r}
obs_data_df <- as_tibbles(snapshot, "observed_data")
head(obs_data_df)
```

## Legacy per-kind helpers

For backwards compatibility, the eight original `get_*_dfs()` functions (`get_compounds_dfs()`, `get_individuals_dfs()`, `get_formulations_dfs()`, `get_populations_dfs()`, `get_events_dfs()`, `get_expression_profiles_dfs()`, `get_protocols_dfs()`, `get_observed_data_dfs()`) remain available as thin wrappers around `as_tibbles()`. New code should prefer `as_tibbles()`.

## Next Steps

- Learn how to create new building blocks in `vignette("creating-building-blocks")`
- Explore the full API in the [function reference](https://esqlabs.github.io/osp.snapshots/reference/)
